Mire microclimate: groundwater buffers temperature in waterlogged versus dry soils

### 2020-10-11 doc/response-reviewers2.Rmd

---
title: "Ref. JOC-20-0599.R1 - Response to reviewers"
author: "Eduardo Fernández-Pascual & Eva Correia-Álvarez"
date: "`r format(Sys.time(), '%d/%m/%Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reviewer: 1

> *My main concern, which I think is possible to smooth out, is linked to the variable length of the monitored time. If there is a comparison made among the sites, this should be based on the same time intervals monitored. In the line 181, I understood that the average for the whole period was calculated. However, the periods were not the same and importantly there are differences among years and an obvious pattern in the data. Looking at Figure 2, there is a clear difference between 2015-2016 and 2019-2020 measurements from any sites (La Recoleta, La Bruxa). It is not clear how much noise this can introduce to the data and results.	I would suggest rerunning the analyses based on data from 2015-2016 only, and for 7 sites which all have the data from this period. If there is no difference, make these results available at least in Supplementary and comment on them in the text. You can also present the data as they are and base your statistical testing only on these data, as being fully comparable.*

To check the reviewer's concern, we have repeated the analysis with only the contemporaneous full-year data series (i.e. Jan 2015-Dec 2017, excluding La Vega Comeya). The output of this analysis did not differ in any significant way from the original analysis using the full datasets. We comment on this in the methods. We have also prepared a supplementary material with the results of the repeated analyses, but in our opinion this supplement adds unnecessary complexity and could be excluded from the final publication, if the editor agrees.

> *Speaking about the testing, please make sure that the data have normal distribution if T-test is your stat (it is sensitive on outliers, especially if n=8). This should be added in the Methods and clearly stated that there was a normal error structure. Otherwise, you better using nonparametric stat as Wilcoxon pair test. The same for Pearson, but the point distribution is shown in Fig. 4 suggests it is okay.*

We have checked whether data meet the normality assumption using the Shapiro-Wilk Normality Test, the assumption was met in all cases. This is now indicated in the text.

> *I am not sure if the statements provided in the lines 208-210 about the wider range and higher maximums were validated based on any stats.*

These statements are validated by the t-tests, described in following paragraphs.

> *I am not sure if the explanation given at 255-259 does not have a flaw in its logic. Maybe I did not get it right, but for me, there is a difference between a gradient and seasonal or diurnal variation. In this part, you are mixing to levels of variation – among sites (differences between sites – a climatic gradient) and within a site, variation over the season or day at the sites. Maybe the problem is only in the wording “the cold end of the thermal gradient”. It confuses me if you call winter and night as the cold end of the gradient. Would it be possible to rephrase this?*

We have substituted the word *gradient* with the word *spectrum*.

> *As there are only 8 values behind each bar in Fig. 3, it might have the sense to present the values themselves. You can then connect individual pairs by a line. Please also add significant above each comparison.*

The change proposed would make Table 2 redundant, and we would prefer to keep it, showing the actual numbers. Since these are the effect sizes in ºC, looking at them gives a quick impression of the strength of the buffer at each site and for each bioclimatic variable.

> *I assume that Patrick A Raney or Patrick A. Raney should be referred from the text as Patrick. It appears at many places in the text, as likely an error of the reference manager.*

Indeed, it was a problem with the reference manager, and should be corrected now.

> *Line 271: missing space at 15 °C (15°C)*

Corrected.

# Reviewer: 2

> *Line 108 – wouldn’t it depend on whether the region is becoming wetter or drier the amount of water being delivered to fens whether they were vulnerable to drying?*

We have changed the sentence to *Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires*.

> *Lines 173-177 –I’d recommend deleting line 173-176 removing the temporary link for peer review to save space in the final manuscript. You can also delete the git-hub reference as you already provide data availability in lines 329.*

We removed these lines.

> *Line 304: micro-refugees  has an extra ‘e’ that should be deleted.*

Corrected.

### 2020-10-11 doc/manuscript.Rmd
minor revision for IJOC

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936]. Both bogs and fens are actively peat forming and are dependent on precipitation, whereas fens rely also on sources of telluric water from mineral ground. Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness. Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935]. Global warming would also result in warmer groundwater delivered to fens and in drier conditions, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].

As is the case for all azonal habitats, locally measured temperatures are essential to understand these processes. Available references recorded temperatures only on waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year, lacking representation throughout the growth cycle of mire vegetation [@RN3204; @RN3060]. This article provides the first measurement of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years. These measures can be used to support previous evidence about the general effects of soil moisture on thermal buffering [@RN3204; @RN2356; @RN4675; @RN4936] and to determine the magnitude of this buffering within microrefugia habitats.  Specifically, we tested the hypotheses that, when compared to adjacent dry soils, waterlogged mire soil are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales. In addition, we compared *in situ* measurements with data derived from the CHELSA climatic models [@RN4766].

@@

This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (Fig. \@ref(fig:fig1)a). Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds. In the study area, average annual precipitation ranges from 800 mm at the low elevations to 1800 mm at the high mountains, while average annual temperature ranges from 12.5 ºC to 5 ºC (source = Agencia Estatal de Meteorología, http://www.aemet.es/es/serviciosclimaticos/). Under the most extreme emissions scenario, the latest projections expect temperatures to increase by 4 ºC and precipitation to decrease by c. 10 % by the end of the century, although the precipitation projections are subjected to great uncertainty [@RN4939]. The present humid and mild climate of this region harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246]. Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on raised hummocks of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes. Glacial lakes undergoing silting develop transition mires and quaking bogs communities (https://eunis.eea.europa.eu/habitats/1637) in the water-to-land transition. Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens (https://eunis.eea.europa.eu/habitats/279) on acid bedrocks, to alkaline calcareous fens (https://eunis.eea.europa.eu/habitats/277) on limestone.

@@

All data processing and analysis was performed in R [@RN2315]. The original data, R code for the analysis and creation of the manuscript can be accessed at the GitHub repository https://github.com/efernandezpascual/mires. To clean the logs we took the following steps: (i) removing records from the first week after installation, to account for the installation process and the settling of the soils; and (ii) keeping only time series with records for both the dry and waterlogged points, because some of the dataloggers had failed at different points in time. Afterwards, we calculated four bioclimatic variables for each datalogger: (1) the mean diurnal range; i.e. the average for the whole period of the daily differences between the maximum and the minimum temperatures recorded in the day; (2) the maximum temperature of the warmest month; i.e. the average of the daily maximum records, for the warmest month; (3) the minimum temperature of the coldest month; i.e. the average of the daily minimum records, for the coldest month; and (4) the annual range; i.e. the difference between the maximum temperature of the warmest month and the minimum temperature of the coldest month. These are classical bioclimatic variables that indicate the limiting factors (extreme temperatures) and seasonality (diurnal and annual range) that organisms must tolerate [@RN4943]. To compare our measurements with model-based predictions of air temperature, we downloaded from CHELSA the same bioclimatic variables for our measuring coordinates (https://chelsa-climate.org/bioclim/).

To test if the differences between dry and waterlogged points were significant, we used paired t-tests (n = 8 paired sites per each of the four bioclimatic variables). Tests were one-tailed, according to the original hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range. To compare our soil measurements with the CHELSA air temperatures we calculated Pearson's correlation between both sets of values, and we also calculated the root-mean-square error (RMSE) of the CHELSA values as a measure of their accuracy (n = 8 sites per combination of bioclimatic variable and dry/waterlogged condition). We calculated RMSE using its implementation in the package *Metrics* [@RN4941].

@@

One important difference with previous studies is the magnitude of the buffer during winter. The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal gradient, i.e. in winter and at night [@RN2356; @RN4675]. In the case of this investigation, the situation is the reverse: the effect is weaker when considering the minimum temperatures of the cold period. This indicates the importance of identifying root temperatures when working with plant communities. At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392]. Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760]. In any case, since this study was conducted at the southern limit of mire distribution, more research on the winter buffer is needed at high latitudes, since it might have a deeper biological significance there.

In these southern mires, the buffering effect is much stronger during the summer (Fig. \@ref(fig:fig5)). This is most noticeable in the two lowest elevations, El Molinucu and La Malva. La Malva shows especially high summer temperatures within the dry soils (> 40 ºC). It must be noted that this difference of > 15ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939]. Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.). This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344]. The buffering is also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e. a former glacial lake undergoing silting. However, the relatively small number of sites in our study does not allow to explore further the moderating factors that may drive differences in the strength of the buffer. More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.

Most models currently employed to predict vegetation and species distribution use macroclimatic parameters, like the ones obtained from CHELSA [@RN4766]. This study revealed that CHELSA values correlate relatively well with soil temperatures in the case of the minimum temperatures of the coldest month, but not with the rest of the bioclimatic variables under examination. Moreover, CHELSA predictions are less accurate for waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil. In addition to the water buffering effect, other features may explain the differences between our results and CHELSEA values, like the height of temperature measurement (5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHELSA); the temporal scale of observation (hourly based for a 5 years period in our study versus a 34 years period); and the spatial resolution (c. 1 m in our study versus c. 1 km). Be that as it may, our results bring out the importance of considering local factors as drivers of microclimatic changes [@RN4762]. These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refugees and buffer species from regional climatic warming.

@@

@@

```{r table3, echo = FALSE}
regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Results of the linear models predicting the soil bioclimatic variables from the CHELSA air temperatures, per bioclimatic variable and groundwater situation. RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).")
```

>>>>

It is evident that the groundwater buffer effect will play a determinant role in the response of mire habitats to climate change [@RN4936]. Both bogs and fens are actively peat forming and are dependent on precipitation, whereas fens rely also on sources of telluric water from mineral ground. Moreover, they show a mosaic of microreliefs and a specific plant cover zonation which makes them especially sensitive to small changes in wetness. Climate change affects the quantity, timing and spatial distribution of precipitation, leading to changes in surface wetness which alter the intensity or organic decomposition by disturbing the conditions for plant grow and the depth of air penetration [@RN4935]. Global warming would also result in warmer groundwater delivered to fens, but there is a gap of knowledge about the rate of such changes and their ecological consequences within the mires [@RN4936].

As is the case for all azonal habitats, locally measured temperatures are essential to understand these processes. Available references recorded temperatures only on waterlogged soils, using model-derived air temperatures for comparison [@RN2356; @RN4675; @RN4679]; or recorded temperatures in both wet and dry spots but for less than a year, lacking representation throughout the growth cycle of mire vegetation [@RN3204; @RN3060]. This article provides the first measurement of the thermal buffer against surrounding non-mire areas, based on soil temperatures recorded during a period of five years. These measures can be used to support previous evidence about the general effects of soil moisture on thermal buffering [@RN3204; @RN2356; @RN4675; @RN4936] and to determine the magnitude of this buffering within microrefugia habitats.  Specifically, we tested the hypotheses that, when compared to adjacent dry soils, waterlogged mire soils are (i) warmer in winter and (ii) colder in summer; and have less thermal amplitude in (iii) daily and (iv) annual scales. In addition, we compared *in situ* measurements with data derived from the CHELSA climatic models [@RN4766].

@@

This study took place in the temperate oceanic region of north-western Spain (43º N, 5º W) (Fig. \@ref(fig:fig1)a). Local geography and climate are dominated by the Cantabrian Mountains (> 1500 m above sea level), which run parallel to the coast and trap the prevailing NW Atlantic winds. In the study area, average annual precipitation ranges from 800 mm at the low elevations to 1800 mm at the high mountains, while average annual temperature ranges from 12.5 ºC to 5 ºC (source = Agencia Estatal de Meteorología, http://www.aemet.es/es/serviciosclimaticos/). Under the most extreme emissions scenario, the latest projections expect temperatures to increase by 4 ºC and precipitation to decrease by c. 10 % by the end of the century, although the precipitation projections are subjected to great uncertainty [@RN4939]. The present humid and mild climate of this region harbours the south-western limit of mire communities in Europe [@RN2982; @RN3246]. Rain-fed raised bogs (https://eunis.eea.europa.eu/habitats/260) are very rare and appear locally on raised hummocks of acid valley mires (https://eunis.eea.europa.eu/habitats/526) which can be found from the coast to just below the treeline, in poorly drained valleys and former glacial lakes. Glacial lakes undergoing silting develop transition mires and quaking bogs communities (https://eunis.eea.europa.eu/habitats/1637) in the water-to-land transition. Spring fens appear in the mountains above 1000 m; they range from soft-water poor fens (https://eunis.eea.europa.eu/habitats/279) on acid bedrocks, to alkaline calcareous fens (https://eunis.eea.europa.eu/habitats/277) on limestone.

@@

All data processing and analysis was performed in R [@RN2315]. To clean the logs we took the following steps: (i) removing records from the first week after installation, to account for the installation process and the settling of the soils; and (ii) keeping only time series with records for both the dry and waterlogged points, because some of the dataloggers had failed at different points in time. Afterwards, we calculated four bioclimatic variables for each datalogger: (1) the mean diurnal range; i.e. the average for the whole period of the daily differences between the maximum and the minimum temperatures recorded in the day; (2) the maximum temperature of the warmest month; i.e. the average of the daily maximum records, for the warmest month; (3) the minimum temperature of the coldest month; i.e. the average of the daily minimum records, for the coldest month; and (4) the annual range; i.e. the difference between the maximum temperature of the warmest month and the minimum temperature of the coldest month. These are classical bioclimatic variables that indicate the limiting factors (extreme temperatures) and seasonality (diurnal and annual range) that organisms must tolerate [@RN4943]. To compare our measurements with model-based predictions of air temperature, we downloaded from CHELSA the same bioclimatic variables for our measuring coordinates (https://chelsa-climate.org/bioclim/).

To test if the differences between dry and waterlogged points were significant, we used paired t-tests (n = 8 paired sites per each of the four bioclimatic variables). Tests were one-tailed, according to the original hypotheses: the dry point would have a higher diurnal range, a higher maximum temperature, a lower minimum temperature, and a higher annual range. To compare our soil measurements with the CHELSA air temperatures we calculated Pearson's correlation between both sets of values, and we also calculated the root-mean-square error (RMSE) of the CHELSA values as a measure of their accuracy (n = 8 sites per combination of bioclimatic variable and dry/waterlogged condition). We calculated RMSE using its implementation in the package *Metrics* [@RN4941]. We checked whether our data complied with the normality assumption of the t-test and Pearson's correlation using the Shapiro-Wilk normality test; the assumption was met in all cases (p > 0.05). To check whether the analyses were influenced by the fact that recording periods did not match across all eight sites, we repeated the analyses using only full-year, contemporaneous data series (01 Jan 2015 - 31 Dec 2017, excluding La Vega Comeya); the results of these analyses were the same as the ones presented in this article (Supplementary Tables S1 and S2).

@@

One important difference with previous studies is the magnitude of the buffer during winter. The articles that had used model air temperatures as a control concluded that the buffer effect was stronger at the cold end of the thermal spectrum, i.e. in winter and at night [@RN2356; @RN4675]. In the case of this investigation, the situation is the reverse: the effect is weaker when considering the minimum temperatures of the cold period. This indicates the importance of identifying root temperatures when working with plant communities. At high elevations of the study region, the soil can remain covered by snow for periods of winter, and this has its own insulating effect on soil temperatures [@RN2392]. Indeed, snow cover has being described as one of the vertical features that affects vegetation distribution in a local manner [@RN4760]. In any case, since this study was conducted at the southern limit of mire distribution, more research on the winter buffer is needed at high latitudes, since it might have a deeper biological significance there.

In these southern mires, the buffering effect is much stronger during the summer (Fig. \@ref(fig:fig5)). This is most noticeable in the two lowest elevations, El Molinucu and La Malva. La Malva shows especially high summer temperatures within the dry soils (> 40 ºC). It must be noted that this difference of > 15 ºC between the dry and waterlogged points of La Malva, which are separated by a few meters, is almost four times the warming expected in the study region at the end of this century (+ 4 ºC) under the more extreme emissions scenario [@RN4939]. Whereas the rest of the study sites are flat, La Malva is a calcareous spring in a southeast-facing slope on limestone, a place experiencing sub-Mediterranean conditions at the micro-scale [@RN4681], and surrounded by a forest of evergreen oaks (*Quercus rotundifolia* Lam., *Quercus faginea* Lam.). This suggests the importance of groundwater in the existence of mire vegetation in Mediterranean areas [@RN4682], not only from the water-availability aspect, but also providing cooling regulation during summer [@RN3344]. The buffering is also prominent at one of the highest elevations, La Recoleta, which may be explained by the high saturation of water in this kind of habitat, a quaking bog, i.e. a former glacial lake undergoing silting. However, the relatively small number of sites in our study does not allow to explore further the moderating factors that may drive differences in the strength of the buffer. More studies focusing on each kind of habitat are needed to understand the relation between hydrology conditions and microclimatic effects.

Most models currently employed to predict vegetation and species distribution use macroclimatic parameters, like the ones obtained from CHELSA [@RN4766]. This study revealed that CHELSA values correlate relatively well with soil temperatures in the case of the minimum temperatures of the coldest month, but not with the rest of the bioclimatic variables under examination. Moreover, CHELSA predictions are less accurate for waterlogged soils, except for the summer max, which might be the consequence of the noise introduced by sun-heated soil. In addition to the water buffering effect, other features may explain the differences between our results and CHELSEA values, like the height of temperature measurement (5 cm below the upper layer of the soil in our study versus 2 m above the ground in CHELSA); the temporal scale of observation (hourly based for a 5 years period in our study versus a 34 years period); and the spatial resolution (c. 1 m in our study versus c. 1 km). Be that as it may, our results bring out the importance of considering local factors as drivers of microclimatic changes [@RN4762]. These factors contribute to landscape heterogeneity, producing safe sites that can act as micro-refuges and buffer species from regional climatic warming.

@@

# Supporting information

*Tables S1-S2* Results of the statistical analyses, repeated using only the contemporaneous full-year data series.

@@

```{r table3, echo = FALSE}
regulartable(table3) %>% width(width = 1) %>% set_caption("Table 3: Comparison of soil bioclimatic variables with CHELSA air temperatures, per bioclimatic variable and groundwater situation. RMSE is the root-mean-square error, a measure of the accuracy of the predictions (lower values indicate higher accuracy).")
```
